Granary weevil eats binary Linux kernels for breakfast

### 2013-10-27 VEE-14/paper.tex
Minor modifications to start things off.

%make title bold and 14 pt font (Latex default is non-bold, 16 pt)
\title{Granary: A Sane Framework for Instrumenting an Insane Environment}

%for single author (just remove % characters)
\authorinfo{Peter Goodman \and Akshay Kumar \and Angela Demke Brown \and Ashvin Goel}
{University of Toronto}{}
% end authorinfo

\maketitle
\subsection*{Abstract}
Kernel modules extend the functionality of operating systems. They represent the bulk of new kernel code in development and contain the majority of new kernel bugs. Unfortunately, analysing and debugging modules is very challenging. Static analysis of module source code is difficult because of the tight interaction between modules and the kernel. Some modules, however, are only distributed in a binary format, which makes static analysis intractable. Existing dynamic analysis tools capable of instrumenting modules are not suited toward module analysis. They either impose unnecessary performance overheads, lack flexibility and are too coarse-grained, or do not provide a means of understanding the ``big picture'' view of what an arbitrary module does as it executes.

We created Granary to address the challenges of module analysis. Granary is a framework for creating flexible and efficient tools that analyse or debug arbitrary, binary Linux kernel modules. Granary uses dynamic binary translation to dynamically rewrite and comprehensively instrument kernel modules. Granary makes it easy to create flexible tools by supporting context-aware runtime code specialisation, and by integrating high-level static analysis information with low-level instruction manipulation. Module analysis tools built on Granary are efficient because of Granary's adoption of a relaxed transparency model, and because of Granary's ability to instrument module code without imposing overhead on non-module kernel code.

@@

Operating system (OS) kernels present an apparently insane environment for a transparent binary instrumentation framework.  For example, hardware interrupts can redirect control flow at any instruction, and interrupt delivery cannot be delayed past the execution of memory instructions that might alter the interrupt delivery state~\cite{DRK}. Preserving interrupt transparency, however, requires that interrupts be delayed until the end of any injected instrumentation instructions.  Together, these requirements make it very difficult to write instrumentation tools.  The insane environment of OS kernels also creates several opportunities for analysis tools. For example, interrupts can be ``tamed'' and put to new uses; privileged hardware features can be used; and the kernel ABI and source code can be relied upon to be a good predictor of module behaviour at the kernel/module boundary. We also found that the kernel is largely insensitive to transparency issues, allowing us to relax transparency in exchange for both improved performance and greater visibility into the execution of instrumented code.  This increased visibility is particularly important for analysis and debugging tools.

The challenges and opportunities of the kernel environment motivated the creation of Granary. Granary is a new binary instrumentation framework for the Linux kernel environment, designed specifically for analysing kernel modules. Granary targets modules because they represent the bulk of new bugs and new code in development in the Linux kernel \cite{FaultsInLinux}. Granary allows developers to create flexible and dynamic analysis tools for arbitrary Linux kernel module binaries. Granary has two main goals: \begin{inparaenum}[i)]
	\item make it easy to create efficient, low-level analysis tools; and
	\item provide high-level static analysis information to the analysis tools.
\end{inparaenum} Granary achieves the first goal by using dynamic binary translation (DBT) to provide low-level access to instructions, memory, and interrupts. Granary achieves the second goal by giving tools direct access to the results of its own static analysis of the Linux kernel source code. By meeting these goals, Granary tools are able to do things like  figure out which data structure fields are associated with each memory access.

Granary addresses three limitations of existing program analysis systems: \begin{inparaenum}[i)]
	\item they have focused on instrumenting all code (the whole kernel: DRK; or whole system: QEMU, PinOS);
	\item they do not allow for flexible runtime code specialisation (Pin, DynamoRIO); and
	\item they do not provide static type information to the instrumented code (Pin, DynamoRIO).
\end{inparaenum}

Instrumenting all code (e.g., the whole kernel) imposes unnecessary performance overheads when only a subset of the code is being analysed (e.g., modules). In practice, we would like to target instrumentation at specific modules for efficiency, while also allowing rich and detailed instrumentation of these modules and any other relevant code. This requirement motivates the design of \emph{mixed-mode execution}, which allows Granary to instrument module code while non-module kernel code executes with zero overhead. This is challenging because Granary must be comprehensive: all code that is meant to be instrumented should be instrumented. In other words, relinquishing control to execute kernel code natively must not cause Granary to miss the execution of any module code.

Another limitation of existing systems is that they are unable to perform efficient runtime code specialisation, i.e. they are unable to change how code is instrumented based on the context in which that code will run. For example, Granary has a tool that  dynamically constructs inter- and intra-procedural control flow graphs (CFGs). The outputs of this tool enable Granary to do better runtime register allocation, which improves tool efficiency. Part of building CFGs requires tracking whether or not a given basic block is the entrypoint to a function. This is challenging with existing DBT systems because some optimised functions will re-execute their first basic block, which might incorrectly appear as new function calls in the CFGs. Granary solves this problem using demand-based runtime code specialisation: every entrypoint basic block is instrumented in one way, and later re-executions of any entrypoint basic blocks are specialised and instrumented in another way.

In Granary, code is specialised according to the context in which that code will execute. In the previous example, executing code is classified into two contexts: entrypoint and non-entrypoint basic blocks. Granary tracks execution contexts using a technique we call \emph{policy-driven instrumentation}. In a Granary tool, each tracked execution context is associated with its own instrumentation policy. An instrumentation policy decides how to instrument code that will execute within the associated context. In the case of our CFG tool, two policies track the two execution contexts: $P_{\mathit{call\_entry}}$ and $P_{\mathit{after\_entry}}$. $P_{\mathit{call\_entry}}$ applies to the first execution of the first basic block of each function, and connects functions in the inter-procedural CFG.  $P_{\mathit{after\_entry}}$ applies to all other executed basic blocks within each function, including re-executions of the first basic block. $P_{\mathit{after\_entry}}$ connects basic blocks in the intra-procedural CFG. Granary is able to track execution contexts because it explicitly maintains policy information in translated basic blocks. This avoids typical problems with tracking context in just-in-time DBT systems (e.g., concurrency, arbitrary pre-emption, context-switch\-ing, inability to predict later execution contexts at translation time).

@@

Granary uses dynamic binary translation (DBT) to instrument Linux kernel modules. DBT is used for instruction set translation/emulation \cite{QEMU}, runtime optimization \cite{DynamoRIOOptimisation}, and runtime instrumentation (analysis \cite{DynamoRIO}, security \cite{Vx32,NaCl}, and debugging \cite{Valgrind}). We chose DBT because \begin{inparaenum}[i)]
	\item static binary analysis tools are unable to cope with \texttt{x86}'s mix of variably-sized instructions and data; 
	\item some modules are only distributed in a binary format; and
	\item virtualisation-based approaches are unable to analyse non-virtualisable device drivers \cite{DRK}.
\end{inparaenum}

Granary interposes on the Linux kernel's module loading process. When a kernel module is loaded, Granary bootstraps by translating the first basic block of the module's initialisation function.  It then replaces the pointer to that function with a pointer to the translated basic block. The translation process continues when the kernel initialises the module by invoking the translated module code.

Granary translates and instruments module binaries one basic block at a time. That is, new basic blocks are decoded and translated as execution ``discovers'' those basic blocks. Translated basic blocks are linked together and stored in a globally accessible \emph{code cache}. In Granary, a basic block is a sequence of instructions ending in a conditional branch, \texttt{ret}, or \texttt{jmp}, but not a \texttt{call} instruction.

@@

Basic blocks in Granary's code cache contain \texttt{x86-64} binary instructions and meta-data describing those instructions. The meta-data that Granary records in each basic block includes: \begin{inparaenum}[i)]
	\item the length in bytes of the original and translated basic blocks;
	\item the address of the first instruction in the original basic block; and
	\item the policy information (e.g., $P_{\mathit{call\_entry}}$, $P_{\mathit{after\_entry}}$) used to instrument the basic block.
\end{inparaenum} This meta-data can be queried and extended by Granary instrumentation tools. For example, Granary's CFG-building tool represents each node in the inter- and intra-procedural CFGs as an extension of the meta-data stored for each basic block. The meta-data is also queried by interrupt handlers when deciding how to handle interrupts in instrumented code, and by debuggers (e.g., \texttt{gdb}) to give contextual information about a translated basic block.

>>>>

%make title bold and 14 pt font (Latex default is non-bold, 16 pt)
\title{Granary weevil eats binary Linux kernels for breakfast}

%for single author (just remove % characters)
\authorinfo{Peter Goodman \and Angela Demke Brown \and Ashvin Goel}
{University of Toronto}{}
% end authorinfo

\maketitle
\subsection*{Abstract}
%Kernel modules extend the functionality of operating systems. They represent the bulk of new kernel code in development and contain the majority of new kernel bugs. Unfortunately, analysing and debugging modules is very challenging. Static analysis of module source code is difficult because of the tight interaction between modules and the kernel. Some modules, however, are only distributed in a binary format, which makes static analysis intractable. Existing dynamic analysis tools capable of instrumenting modules are not suited toward module analysis. They either impose unnecessary performance overheads, lack flexibility and are too coarse-grained, or do not provide a means of understanding the ``big picture'' view of what an arbitrary module does as it executes.

%We created Granary to address the challenges of module analysis. Granary is a framework for creating flexible and efficient tools that analyse or debug arbitrary, binary Linux kernel modules. Granary uses dynamic binary translation to dynamically rewrite and comprehensively instrument kernel modules. Granary makes it easy to create flexible tools by supporting context-aware runtime code specialisation, and by integrating high-level static analysis information with low-level instruction manipulation. Module analysis tools built on Granary are efficient because of Granary's adoption of a relaxed transparency model, and because of Granary's ability to instrument module code without imposing overhead on non-module kernel code.

@@

%Operating system (OS) kernels present an apparently insane environment for a transparent binary instrumentation framework.  For example, hardware interrupts can redirect control flow at any instruction, and interrupt delivery cannot be delayed past the execution of memory instructions that might alter the interrupt delivery state~\cite{DRK}. Preserving interrupt transparency, however, requires that interrupts be delayed until the end of any injected instrumentation instructions.  Together, these requirements make it very difficult to write instrumentation tools.  The insane environment of OS kernels also creates several opportunities for analysis tools. For example, interrupts can be ``tamed'' and put to new uses; privileged hardware features can be used; and the kernel ABI and source code can be relied upon to be a good predictor of module behaviour at the kernel/module boundary. We also found that the kernel is largely insensitive to transparency issues, allowing us to relax transparency in exchange for both improved performance and greater visibility into the execution of instrumented code.  This increased visibility is particularly important for analysis and debugging tools.

%The challenges and opportunities of the kernel environment motivated the creation of Granary. Granary is a new binary instrumentation framework for the Linux kernel environment, designed specifically for analysing kernel modules. Granary targets modules because they represent the bulk of new bugs and new code in development in the Linux kernel \cite{FaultsInLinux}. Granary allows developers to create flexible and dynamic analysis tools for arbitrary Linux kernel module binaries. Granary has two main goals: \begin{inparaenum}[i)]
%	\item make it easy to create efficient, low-level analysis tools; and
%	\item provide high-level static analysis information to the analysis tools.
%\end{inparaenum} Granary achieves the first goal by using dynamic binary translation (DBT) to provide low-level access to instructions, memory, and interrupts. Granary achieves the second goal by giving tools direct access to the results of its own static analysis of the Linux kernel source code. By meeting these goals, Granary tools are able to do things like  figure out which data structure fields are associated with each memory access.

Granary addresses three limitations of existing program analysis systems: \begin{inparaenum}[i)]
	\item they have focused on instrumenting all code (the whole kernel: DRK, btkernel; or whole system: QEMU, PinOS);
	\item they do not allow for flexible runtime code specialisation; and
	\item they do not provide static type information to the instrumented code.
\end{inparaenum}

Instrumenting all code (e.g., the whole kernel) imposes unnecessary performance overheads when only a subset of the code is being analysed (e.g., modules, functions). In practice, we would like to target instrumentation at specific code for efficiency, while also allowing rich and detailed instrumentation of the targeted code. This requirement motivates the design of \emph{mixed-mode execution} and \emph{probe-based instrumentation}, which allows Granary to instrument specific module code or functions while the majority of kernel code executes with zero overhead.

%This is challenging because Granary must be comprehensive: all code that is meant to be instrumented should be instrumented. In other words, relinquishing control to execute kernel code natively must not cause Granary to miss the execution of any module code.

Another limitation of existing systems is that they are unable to perform efficient runtime code specialisation, i.e. they are unable to change how code is instrumented based on the context in which that code will run.

%For example, Granary has a tool that  dynamically constructs inter- and intra-procedural control flow graphs (CFGs). The outputs of this tool enable Granary to do better runtime register allocation, which improves tool efficiency. Part of building CFGs requires tracking whether or not a given basic block is the entrypoint to a function. This is challenging with existing DBT systems because some optimised functions will re-execute their first basic block, which might incorrectly appear as new function calls in the CFGs. Granary solves this problem using demand-based runtime code specialisation: every entrypoint basic block is instrumented in one way, and later re-executions of any entrypoint basic blocks are specialised and instrumented in another way.

In Granary, code is specialised according to the context in which that code will execute. Granary tracks execution contexts using a technique we call \emph{policy-driven instrumentation}. In a Granary tool, each tracked execution context is associated with its own instrumentation policy. An instrumentation policy decides how to instrument code that will execute within the associated context. Granary is able to track execution contexts because it explicitly maintains policy information in translated basic blocks. This avoids typical problems with tracking context in just-in-time DBT systems (e.g., concurrency, arbitrary pre-emption, context-switch\-ing, inability to predict later execution contexts at translation time).

% In the previous example, executing code is classified into two contexts: entrypoint and non-entrypoint basic blocks.  In the case of our CFG tool, two policies track the two execution contexts: $P_{\mathit{call\_entry}}$ and $P_{\mathit{after\_entry}}$. $P_{\mathit{call\_entry}}$ applies to the first execution of the first basic block of each function, and connects functions in the inter-procedural CFG.  $P_{\mathit{after\_entry}}$ applies to all other executed basic blocks within each function, including re-executions of the first basic block. $P_{\mathit{after\_entry}}$ connects basic blocks in the intra-procedural CFG.

@@

Granary uses dynamic binary translation (DBT) to instrument the Linux kernel. DBT is used for instruction set translation/emulation \cite{QEMU}, runtime optimization \cite{DynamoRIOOptimisation}, and runtime instrumentation (analysis \cite{DynamoRIO}, security \cite{Vx32,NaCl}, and debugging \cite{Valgrind}). We chose DBT because \begin{inparaenum}[i)]
	\item static binary analysis tools are unable to cope with \texttt{x86}'s mix of variably-sized instructions and data; 
	\item some kernel code is only distributed in a binary format (e.g. proprietary device drivers); and
	\item virtualisation-based approaches are unable to analyse non-virtualisable device drivers \cite{DRK}.
\end{inparaenum}

Granary instruments code according to one of three strategies: \begin{inparaenum}[i)]
	\item full-kernel instrumentation;
	\item module-only instrumentation, and;
	\item probe-based instrumentation.
\end{inparaenum}

\paragraph{Full-kernel instrumentation}
Granary interposes on the system call and interrupt entrypoints, located in the x86-64 \texttt{MSR\_LSTAR} model-specific register and interrupt descriptor table (IDT), respectively. Granary boostraps by duplicating the early entrypoint instructions up until the kernel's native entrypoint switches from a user-space to a kernel-space stack. Granary translates the first basic block immediately following the stack switch, and adds a control-flow transfer instruction from the duplicated entrypoint instructions into the translated basic block. The translation process continues when a system call or interrupt occurs. Our choice to defer translation to a point where execution operates off of a kernel stack simplifies instrumentation tools that want to save and restore transient data on the runtime call stack.

\paragraph{Module-only instrumentation}
Granary interposes on the Linux kernel's module loading process. When a kernel module is loaded, Granary bootstraps by translating the first basic block of the module's initialisation function.  It then replaces the pointer to that function with a pointer to the translated basic block. The translation process continues when the kernel initialises the module by invoking the translated module code.

\paragraph{Probe-based instrumentation}
Granary can interpose on the execution of specific Linux kernel functions in three ways:
\begin{enumerate}
	\item {\bf Wrapping:} Granary can substitute the execution of a kernel function when that function is invoked from instrumented code. A different function can be substituted depending on the context from which the original function is invoked (instrumented kernel code, instrumented module code). We call this type of interposition ``wrapping'' because it is commonly used to inspect the substituted function's arguments and return value, while still invoking the original function.
	\item {\bf Substituting:}
	\item {\bf Instrumented Wrapping/Substituting:}
\end{enumerate}

Granary translates and instruments kernel binaries one basic block at a time. That is, new basic blocks are decoded and translated as execution ``discovers'' those basic blocks. Translated basic blocks are linked together and stored in a globally accessible \emph{code cache}. In Granary, a basic block is a sequence of instructions ending in a conditional branch, \texttt{ret}, or \texttt{jmp}, but not a \texttt{call} instruction.

@@

Basic blocks in Granary's code cache contain x86-64 binary instructions and meta-data describing those instructions. The meta-data that Granary records in each basic block includes: \begin{inparaenum}[i)]
	\item the length in bytes of the original and translated basic blocks;
	\item the address of the first instruction in the original basic block; and
	\item the policy information (e.g., $P_{\mathit{call\_entry}}$, $P_{\mathit{after\_entry}}$) used to instrument the basic block.
\end{inparaenum} This meta-data can be queried and extended by Granary instrumentation tools. For example, Granary's CFG-building tool represents each node in the inter- and intra-procedural CFGs as an extension of the meta-data stored for each basic block. The meta-data is also queried by interrupt handlers when deciding how to handle interrupts in instrumented code, and by debuggers (e.g., \texttt{gdb}) to give contextual information about a translated basic block.
